Ionization of biological molecules by multicharged ions by using the stoichiometric model

### 2019-09-23 NewMonti25.tex
claudia's comments included

%\documentclass[pra,showpacs,twocolumn]{revtex4}

\documentclass[preprint,12pt]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{gensymb}
\usepackage{float}
\usepackage{authblk}
\usepackage{MnSymbol}

\setcounter{MaxMatrixCols}{10}
%TCIDATA{OutputFilter=LATEX.DLL}
%TCIDATA{Version=5.50.0.2953}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{BibliographyScheme=Manual}
%TCIDATA{LastRevised=Thursday, March 07, 2019 14:11:39}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{Language=American English}

%\input{tcilatex}
\begin{document}

\title{Ionization of biological molecules by multicharged ions by using 
the stoichiometric model}
\author{author}
\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\author{author}
%\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\affil{Fac. de Ciencias Exactas y Naturales, Universidad de Buenos Aires}
\date{\today }

@@

%\pacs{79.20.Rf, 68.49.Bc, 34.20.Cf,34.50.-s,34.35.+a}

@@

The damage caused by the impact of multicharged heavy projectiles on 
biological targets has become a field of interest due to its recent 
implementation in ion--beam cancer therapy. The effectiveness of the 
radiation depends on the choice of the ions. In particular, theoretical 
and experimental studies with different projectiles have concluded that 
charged carbon ions could be the most suitable ions to be used. The 
study of such systems represent a challenge from the theoretical point 
of view. The most widely used method to compute ionization of 
multicharged atoms is the first Born approximation. This perturbative 
method warrants the $Z^{2}$ laws, where $Z$ is the projectile charge. 
However, the damage is concentrated in the vicinities of the Bragg peak,
at energies of hundreds of keV/amu. Precisely in this intermediate 
region, the Born approximation starts to fail. Another theoretical 
issue arises from the targets themselves; we are dealing with complex 
molecules, which are a hard task for {\it ab initio} calculations. 
The objective of this article is to deal with these two aspects. 
First, we perform more appropriate calculations on the primary damage 
mechanism, i.e., atomic ionization by multicharged ions, which can 
replace the Born results. Second, we inspect and use an stoichiometric 
model, which reduces a molecule to a sum of atomic processes quantities 
weighted by the number of such atom in the molecule.

To overcome the first perturbative approximation limitations, and 
since the projectiles are multicharged ions, we resort to the 
Continuum Distorted Wave-Eikonal Initial State (CDW), which includes 
higher perturbative corrections. We start from the premise that the 
ionization process is the mechanism that deposits the most significant 
amount of primary energy. This process produces an electron-energy 
spectrum that needs to be integrated. Moreover, the ejected electrons 
become a new source of local damage. These secondary electrons are 
included in Monte Carlo simulations and hence their behavior must be
investigated. With this end in mind, we calculate not only the 
ionization cross-sections but also energy and mean angular distributions 
of the emitted electrons.

The molecular structure complexity of the target is dealt with by 
implementing the simplest stoichiometric model (SSM); we assume that 
the molecule is composed of isolated independent atoms. In all cases, 
the total cross-section is simply the sum of the cross-section of each 
atom.

Then, by implementing the SSM and the CDW --instead of the first Born 
approximation--, we calculate ionization cross-section of several 
molecules (see Table~\ref{tab:families}) by the impact of antiprotons,
H$^{+}$, He$^{+2}$, Be$^{+4}$, C$^{+6}$,\ and O$^{+8}$. Furthermore, 
in Section 3, we investigate different DNA and RNA molecules
such as adenine, cytosine, guanine, thymine, uracil, tetrahydrofuran 
(THF), pyrimidine and also DNA backbone.

The results are processed to test the Toburen scaling rule, which 
states that the ratio between the ionization cross-section and the 
number of valence electrons (outer electrons) in terms of the 
projectile velocity can be arranged in a narrow universal band. 
We have also applied this rule to several hydrocarbons and nucleobases. 
In Section 3.2, we will prove that the width of the resulting universal 
band can be significantly reduced if we redefine the effective number 
of valence electrons.

It is well known that the residual electrons from the ionization 
process cause significant biological damage. To inspect this mechanism, 
in Section 3.3 and 3.4, we calculated the mean electron energy and
angular distributions. Surprisingly, we found a substantial dependence 
of the charged projectile, which is unexpected in the first Born 
approximation.

The hefty charge projectiles are dealt with the CDW. However, the 
stoichiometric model used seems to be very simplistic. The approximation
considers each atom as neutral, which is not correct. In Section 3.5, 
we used the molecular electronic structure code 
{\sc gamess}~\cite{gamess} to calculate the excess or defect of 
electron density one each atom. Then, the simple stoichiometric model is
modified to account for the departure from neutrality of the atoms. 
We find that for the DNA molecules this modification does not introduce 
a substantial change.

@@

In our study, we will consider six atoms $\alpha=$ H, C, N, O, P, and S. 
Most of the organic molecules are composed of these atoms. Some 
particular molecules also include halogen atoms such as fluor and 
bromine; ionization cross-sections of these elements have been 
previously published~\cite{miraglia2008,miraglia2009}.

The total ionization cross-sections of these atoms were calculated using 
the CDW. The initial bound and final continuum radial wave functions 
were obtained by using the {\sc radialf} code, developed by Salvat and 
co-workers, and a Hartree-Fock potential obtained from the Depurated 
Inversion Model~\cite{mendez2016,mendez2018}. 
We used a few thousand pivot points to solve the Schr\"{o}dinger 
equation, depending on the number of oscillations of the continuum 
state. The radial integration was performed using the 
cubic spline technique. The number of angular momenta considered 
varied from 8, at very low ejected-electron energies, up to 30, 
for the highest energies considered. The same number of azimuth 
angles were required to obtain the four-fold differential 
cross-section. The calculation performed does not display prior-post 
discrepancies at all. Each atomic total cross-section in 
Eq.~(\ref{eq:sumion}) was calculated using 35 to 100 momentum transfer 
values, 28 fixed electron angles, and around 45 electron energies 
depending on the projectile impact energy. Further details of the 
calculation are given in Ref.~\cite{montanari2017}. Simultaneously, we 
will be reporting state to state ionization cross sections for these 
cases in Ref.~\cite{miraglia2019}, which will be very useful to estimate 
molecule fragmentation.

In Fig.~\ref{fig:atomscaling}, we report our total ionization results 
for the six essential elements by the impact of six different projectiles: 
antiprotons, H$^{+}$, He$^{+2}$, Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
To reduce the resulting 36 magnitudes into a single consistent 
figure, we considered the fact that in the first Born approximation
the ionization cross-section scales with the square of the projectile 
charge, $Z^{2}$. The values of the impact energies considered 
range between 0.1 to 10 MeV/amu, where the CDW is supposed 
to hold. In fact, for the highest projectile charges the minimum 
impact energy where the CDW is expected to be valid could be 
higher than 100 KeV. Moreover, we corroborated that the first Born 
approximation provides quite reliable results only for energies higher 
than a couple of MeV/amu. We use the same color to indicate the 
projectile charge throughout the figures of this work: dashed-red, 
solid-red, blue, magenta, olive and orange for antiprotons, 
H$^{+}$, He$^{+2}$, Be$^{+4}$, C$^{+6}$, and O$^{+8}$, respectively. 
The charge state of the projectiles are also given in the figures.
Notably, there is no complete tabulation of ionization of atoms by 
the impact of multicharged ions. We hope that the ones presented 
in this article will be of help for future works.

\begin{figure}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/Fig1.eps}
\includegraphics[width=\textwidth]{figuras/atomicscaling.eps}
\caption{Scaled total ionization cross section of six atomic targets 
by impact of multicharged projectiles.}
\label{fig:atomscaling}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Emitted electron energies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In a given biological medium, direct ionization by ion impact accounts 
for just a fraction of the overall damage. Secondary electrons, as well 
as recoil target ions, also contribute substantially to the total damage. 
We can consider the single differential cross section of the shell 
$nl$ of the atom $\alpha$, $d\sigma_{\alpha nl}/dE$, to be a function 
of the ejected electron energy $E$ as a simple distribution 
function~\cite{surdutovic2018}. Then, we can define the mean value 
$\overline{E}_{\alpha}$ as in Ref.~\cite{abril2015},
\begin{eqnarray}
\overline{E}_{\alpha} &=&\frac{\langle E_{\alpha}\rangle}{\langle
1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}\int dE\,E
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{40} \\
\langle 1\rangle &=&\sigma_{\alpha}=\sum\limits_{nl}\int dE\,
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{50}
\end{eqnarray}
where $\Sigma_{nl}$ takes into account the sum of the different 
sub-shell contributions of the element $\alpha$.

\begin{figure}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig5.eps}
\includegraphics[width=\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by impact of
multicharged ions. Charge state of projectiles given by arrow.}
\label{fig:emittedener}
\end{figure}

Figure~\ref{fig:emittedener} shows $\overline{E}_{\alpha}$ for six elements from 
Table~\ref{tab:families}. The range of impact velocities was shorten up 
to $v=10$ a.u. due to numerical limitations in the spherical harmonics 
expansion. In our theoretical treatment, we expand our final continuum 
wave function as per usual,
\begin{equation}
\psi_{\overrightarrow{k}}^{-}(\overrightarrow{r})=\sum_{l=0}^{l_{\max
}}\sum_{m=-l}^{l}R_{kl}^{-}(r)Y_{l}^{m}(\widehat{r})Y_{l}^{m^{\ast }}
(\widehat{k})\,.
\label{60}
\end{equation}
We are confident with our calculations up to $l_{\max}\sim 30$. 
As the impact velocity $v$ increases, so do $\langle E_{\alpha}\rangle$
and $l_{\max}$, which results in the inclusion of very oscillatory 
functions in the integrand. Furthermore, the integrand of
$\langle E_{\alpha}\rangle$ includes the kinetic energy $E$
(see Eq.~(\ref{40})), which cancels the small energy region and 
reinforces the large values, making the result more sensible to large
angular momenta. Regardless, for $v>10$ a.u., the first Born 
approximation holds.

In Figure~\ref{fig:emittedener} , we estimate $\overline{E}_{\alpha}$ 
in the 0.5--2 a.u.
velocity range, or equivalently from 15 to 50 eV, for all the targets.
Our results agree with the experimental findings~\cite{surdutovic2018}. 
The dependence of the mean energy value with the projectile charge $Z$ 
is surprisingly sensible, which can duplicate the proton results. 
This effect can be attributed to the depletion caused by the 
multicharged ions to the yields of low energy electrons. In the high 
approximation, surviving the $Z^{2}$ law. Then, the ratio in 
Eq.~(\ref{40}) cancels out and $\overline{E}_{\alpha}$ becomes a 
universal value independent on $Z$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Emitted electron angles}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig6.eps}
\includegraphics[width=\textwidth]{figuras/ang_mean.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure}

Figure~\ref{fig:emittedang} shows $\overline{\theta}_{\alpha}$ for our six elements of 
interest. A new important dependence of $\overline{\theta}_{\alpha}$ 
with $Z$ is observed. It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons are nearly isotropic. This effect 
is caused by the insignificant angular anisotropy of sub-50-eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. But, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region, as observed in Figure~\ref{fig:emittedang}. 
For example, for C$^{+6}$ impact, the Bragg peak occurs at 0.3 MeV/amu, 
where $\overline{\theta}_{\alpha}$, computed with the CDW method, is 
about half of the value obtained with the first Born approximation. 
This correction should close the damage to the forward direction.

We can attribute this correction to the capture to the continuum effect;
the larger the charge $Z$, the smaller $\overline{\theta}$ will be. Of 
course, this effect only holds at intermediate energies and not at high 
impact energies, where the Born approximation rules. One illustrative 
observation is the behavior of antiprotons: the projectile in this case 
repels the electrons, making the distribution almost symmetric. 
Note the opposite effect of proton and antiprotons; they run one 
opposite to the other, as compared with the first Born approximation.

@@

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM describes the total ionization cross section 
of the molecule $\sigma_{M}$ as a simple sum of ionization cross 
sections of the isolated atoms $\sigma_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We divided seventeen molecular targets of our interest in three families. 
The classification defined is given in Table~\ref{tab:families}.

\begin{table}[H]
\begin{center}
\begin{tabular}{|p{0.06\textwidth}|p{0.75\textwidth}|}
\hline
 CH  & CH$_4$ (methane), C$_2$H$_2$ (acetylene), \\
     & C$_2$H$_4$ (ethene), C$_2$H$_6$ (ethane), C$_6$H$_6$ (benzene) \\
\hline
 CHN & C$_5$H$_5$N (pyridine), C$_4$H$_4$N$_2$ (pyrimidine), \\
& C$_2$H$_7$N 
(dimenthylamine), CH$_5$N (monomethylamine) \\
\hline
 \multirow{4}{*}{DNA} 
     & C$_5$H$_5$N$_5$ (adenine), C$_4$H$_5$N$_3$O (cytosine), \\
     & C$_5$H$_5$N$_5$O (guanine), C$_5$H$_6$N$_2$O$_2$ (thymine), \\
     & C$_4$H$_4$N$_2$O$_2$ (uracil), C$_4$H$_8$O (THF), \\
     & C$_5$H$_{10}$O$_5$P (DNA backbone), C$_{20}$H$_{27}$N$_7$O$_{13}$P$_2$ (dry DNA) \\
\hline
\end{tabular}
\caption{Molecular targets of our interest classified in three families.}
\label{tab:families}
\end{center}
\end{table}

We report the total ionization cross sections by the impact of 
multicharged ions for adenine, cytosine, guanine and thymine in 
Fig.~\ref{fig:crossDNA_1}. For adenine, the agreement with the 
experimental data available~\cite{iriki2011} is very good 
in our range of validity. To the best of our knowledge, there are not 
experimental data on ion--collision ionization for the rest of the 
molecules. We included electron impact measurements~\cite{rahman2016} 
with the corresponding equivelocity 
conversion for incident energies higher than 500~keV. In this region, 
the proton and electron cross section should agree. Although the electron 
impact measurements are above our findings for all the molecular targets, 
it is worth stating that our results agree very well with other 
theoretical predictions~\cite{mozejko2003,tan2018}.

\begin{figure}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig2.eps}
\includegraphics[width=\textwidth]{figuras/adn1.eps}
\caption{Scaled ionization cross section by impact of multicharged ions. 
Experiments: 
$\circ$~\cite{iriki2011} for proton impact and $\square$~\cite{rahman2016} 
for electron impact with equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure}

>>>>

%\documentclass[pra,showpacs,twocolumn]{revtex4}
\documentclass[preprint]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{gensymb}
\usepackage{float}
\usepackage{authblk}
\usepackage{MnSymbol}
\usepackage{gensymb}

\begin{document}

\title{Ionization of biological molecules by multicharged ions by using 
the stoichiometric model}

\author{author}
%\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio 
%(CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\author{author}
%\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\affil{Fac. de Ciencias Exactas y Naturales, Universidad de Buenos Aires}

\date{\today}% It is always \today, today,

%\keywords{Suggested keywords}

@@

\tableofcontents
%\pacs{79.20.Rf, 68.49.Bc, 34.20.Cf,34.50.-s,34.35.+a}

@@

The damage caused by the impact of multicharged heavy projectiles on 
biological targets has become a field of interest due to its recent 
implementation in ion--beam cancer therapy. The effectiveness of the 
radiation depends on the choice of the ions. In particular, theoretical 
and experimental studies with different projectiles have concluded that 
charged carbon ions could be the most suitable ions to be used. 
Nonetheless, the study of such systems represents a challenge from the 
theoretical point of view.

The ionization of biological molecules by multicharged ions constitutes 
the primary damage mechanism. The most widely used method to predict
such processes is the first Born approximation. At high energies, this 
perturbative method warrants the $Z^{2}$ laws, where $Z$ is the 
projectile charge. However, the damage is concentrated in the 
vicinities of the Bragg peak --at energies of hundreds of keV/amu--, 
precisely where the Born approximation starts to fail. 
Another theoretical issue arises due to the targets themselves; we are 
dealing with complex molecules, and the description of such targets 
represents a hard task for {\it ab initio} calculations. The objective 
of this article is to deal with these two aspects; first, we perform 
more appropriate calculations on the primary damage mechanism, which can 
replace the Born results. Second, we inspect and test a stoichiometric 
model to describe the ionization of molecular targets.

To overcome the first perturbative approximation limitations, and since 
the projectiles are multicharged ions, we resort to the Continuum 
Distorted Wave--Eikonal Initial State (CDW), which includes higher 
perturbative corrections. We start from the premise that the ionization 
process is the mechanism that deposits the most significant amount of 
primary energy. Moreover, it is well known that the residual electrons 
from the ionization become a source of significant local biological 
damage. The secondary electrons are included in Monte Carlo simulations, 
and hence their behavior must be investigated. 
In Section~\ref{subsec:meanener} and \ref{subsec:meanang}, we calculate
the mean energy and angular distributions of the ejected electrons. 
Surprisingly, we found a substantial dependence of the charged 
projectile, which is unexpected in the first Born approximation. 
We deal with the molecular structure complexity of the target by 
implementing the simplest stoichiometric model (SSM): the molecules are 
assumed to be composed of isolated independent atoms, and the total 
cross--section by a linear combination of weighted atomic calculations.

By implementing the CDW and the SSM, we calculate ionization 
cross--section of several molecules of interest (see 
Table~\ref{tab:families}) by the impact of antiprotons, H$^{+}$, 
He$^{+2}$, Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
In Section~3, we show our results for different DNA and RNA molecules, 
such as adenine, cytosine, guanine, thymine, uracil, tetrahydrofuran~(THF), 
pyrimidine, and also DNA backbone. In Section~\ref{subsec:scaling}, we
test the Toburen scaling rule~\cite{toburen1975,toburen1976}, which 
states that the ratio between the ionization cross--section and the 
number of weakly bound electrons can be arranged in a narrow universal 
band in terms of the projectile velocity. We applied this rule to 
several hydrocarbons and nucleobases and noted that the width of the 
resulting universal band could be significantly reduced if we redefine 
the effective number of active electrons in the collision. The new 
scaling is then tested by comparison with experimental data.

Although the multicharged projectiles are dealt with the CDW, the 
stoichiometric model used seems to be very simplistic. The approach
considers each atom as neutral, which is not correct. In Section 3.5, 
we used the molecular electronic structure code 
{\sc gamess}~\cite{gamess} to calculate the excess or defect of 
electron density on the atoms composing the molecules. Then, the SSM is 
modified to account for the departure from the neutrality of the atoms. 
We find that the modified SSM for the DNA molecules does not introduce 
substantial changes in the cross sections.

@@

In the present study, we consider six atoms, $\alpha=$ H, C, N, O, P, 
and S, and six projectiles, antiprotons $\bar{p}$, H$^{+}$, He$^{+2}$, 
Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
Most of the organic molecules are composed of these atoms. Some 
particular molecules also include halogen atoms such as fluor and 
bromine; ionization cross sections of these elements have been 
previously published~\cite{miraglia2008,miraglia2009}.

The total ionization cross sections of these atoms were calculated using 
the CDW~[xxx]. The initial bound and final continuum radial wave 
functions were obtained by using the {\sc radialf} code, developed by 
Salvat and co--workers~\cite{salvat1995}, and a Hartree-Fock potential 
obtained from the Depurated Inversion Method~\cite{mendez2016,mendez2018}. 
We used a few thousand pivot points to solve the Schr\"{o}dinger 
equation, depending on the number of oscillations of the continuum 
state. The radial integration was performed using the cubic spline 
technique. The number of angular momenta considered varied from 8, at 
very low ejected-electron energies, up to 30, for the highest energies 
considered. The same number of azimuth angles were required to obtain 
the four-fold differential cross section. The calculation performed does 
not display prior-post discrepancies at all. Each atomic total cross 
section was calculated using 35 to 100 momentum transfer values, 28 
fixed electron angles, and around 45 electron energies depending on the 
projectile impact energy. In our theoretical treatment, we expand our 
final continuum wave function as usual,
\begin{equation}
\psi_{\overrightarrow{k}}^{-}(\overrightarrow{r})=\sum_{l=0}^{l_{\max
}}\sum_{m=-l}^{l}R_{kl}^{-}(r)Y_{l}^{m}(\widehat{r})Y_{l}^{m^{\ast }}
(\widehat{k})\,.
\label{eq:contwave}
\end{equation}
We are confident with our calculations up to $l_{\max}\sim 30$. Further 
details of the calculation are given in Ref.~\cite{montanari2017}. 
Simultaneously, we will be reporting state to state ionization cross 
sections for the 36 ion--target systems considered in the present 
work~\cite{miraglia2019}. We expect that these results will be useful
to estimate molecule fragmentation.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/Fig1.eps}
\includegraphics[width=0.9\textwidth]{figuras/atomicscaling.eps}
\caption{Reduced CDW total ionization cross section of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

We report our total CDW ionization cross sections for the six essential 
elements by the impact of six projectiles in Fig.~\ref{fig:atomscaling}.
To reduce the resulting 36 magnitudes into a single consistent 
figure, we considered the fact that in the first Born approximation
the ionization cross section scales with the square of the projectile 
charge, $Z^{2}$. The values of the impact energies considered 
range between 0.1 to 10 MeV/amu, where the CDW is supposed 
to hold. In fact, for the highest projectile charges the minimum 
impact energy where the CDW is expected to be valid could be 
higher than 100 keV. We also performed similar calculations with the 
first Born approximation, and we corroborated that it provides quite 
reliable results only for energies higher than a couple of MeV/amu. 
We use the same line color to indicate the projectile charge throughout 
all the figures of this work: dashed--red, solid--red, blue, magenta, 
olive and orange for antiprotons, H$^{+}$, He$^{+2}$, Be$^{+4}$, 
C$^{+6}$, and O$^{+8}$, respectively. Notably, there is no complete 
tabulation of ionization of atoms by the impact of multicharged ions. 
We hope that the ones presented in this article will be of help for 
future works.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Emitted electron energies}
\label{subsec:meanener}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In a given biological medium, direct ionization by ion impact accounts 
for just a fraction of the overall damage. Secondary electrons, as well 
as recoil target ions, also contribute substantially to the total damage. 
We can consider the single differential cross section of the shell 
$nl$ of the atom $\alpha$, $d\sigma_{\alpha nl}/dE$, to be a function 
of the ejected electron energy $E$ as a simple distribution 
function~\cite{surdutovic2018}. Then, we can define the mean value 
$\overline{E}_{\alpha}$ as in Ref.~\cite{abril2015},
\begin{eqnarray}
\overline{E}_{\alpha} &=&\frac{\langle E_{\alpha}\rangle}{\langle
1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}\int dE\,E
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{40} \\
\langle 1\rangle &=&\sigma_{\alpha}=\sum\limits_{nl}\int dE\,
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{50}
\end{eqnarray}
where $\Sigma_{nl}$ takes into account the sum of the different 
sub--shell contributions of the element $\alpha$.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig5.eps}
\includegraphics[width=0.9\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by impact of
multicharged ions, given by Eq.~(\ref{40}). Dashed lines for $\bar{p}$,
solid lines for ion charges $+1$, $+2$, $+4$, $+6$ and $+8$,
as indicated.}
\label{fig:emittedener}
\end{figure*}

The mean emitted electron energies $\overline{E}_{\alpha}$ for H, C, N, 
O, P and S are shown in Fig.~\ref{fig:emittedener}. The range of impact 
velocities was shortened to $v=10$ a.u. due to numerical limitations 
in the spherical harmonics expansion of Eq.~(\ref{eq:contwave}). 
As the impact velocity $v$ increases, so do $\langle E_{\alpha}\rangle$
and $l_{\max}$, which results in the inclusion of very oscillatory 
functions in the integrand. Furthermore, the integrand of
$\langle E_{\alpha}\rangle$ includes the kinetic energy $E$
(see Eq.~(\ref{40})), which cancels the small energy region and 
reinforces the large values, making the result more sensible to large
angular momenta. Regardless, for $v>10$ a.u., the first Born 
approximation holds.

In Fig.~\ref{fig:emittedener}, we estimate $\overline{E}_{\alpha}$ of
the emitted electron in the 0.5--2.7 a.u. energy range, or equivalently 
from 15 to 70 eV, for all the targets. Our results agree with the 
experimental findings~\cite{surdutovic2018}. As can be noted in the 
figure, the mean energy value is surprisingly sensible to the 
projectile charge $Z$, which can duplicate the proton results in the 
intermediate region, i.e. 100--400 keV/amu. The effect observed can be 
attributed to the depletion caused by the multicharged ions to the 
yields of low energy electrons. This behavior cannot be found in the 
first Born approximation, where the $Z^2$ law cancels the $Z$ dependence
in Eq.~(\ref{40}). At high energies, $\overline{E}_{\alpha}$ tends to a 
universal value for all ions, as can be seen in Fig.~\ref{fig:emittedener}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Emitted electron angles}
\label{subsec:meanang}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig6.eps}
\includegraphics[width=0.9\textwidth]{figuras/ang_mean.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure*}

The mean emitted electron angles $\overline{\theta}_{\alpha}$ for the 
six atoms of interest are shown in Fig.~\ref{fig:emittedang}, and an 
important dependence of $\overline{\theta}_{\alpha}$ with $Z$ is 
observed. Once again, this effect is not observed in the first Born 
approximation. It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons are nearly isotropic. This 
behavior is caused by the insignificant angular anisotropy of 
sub--50--eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. But, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region, as observed in Fig.~\ref{fig:emittedang}. 
For example, for C$^{+6}$ impact, the Bragg peak occurs at 0.3 MeV/amu, 
where $\overline{\theta}_{\alpha}$, computed with the CDW method, is 
about half of the value obtained with the first Born approximation. 
This correction closes the damage to the forward direction.

We can attribute this forward direction correction to the capture to 
the continuum effect; the larger the charge $Z$, the smaller 
$\overline{\theta}$ will be. Of course, this effect only holds at 
intermediate energies and not at high impact energies, where the Born 
approximation rules. One illustrative observation is the behavior of 
antiprotons: the projectile in this case repels the electrons, being
$\overline{\theta}_{\alpha}\sim 90\degree$.
Note the opposite effect of proton and antiprotons; they run one 
opposite to the other, as compared with the first Born approximation,
constituting an angular Barkas effect.

@@

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM approaches the total ionization cross section 
of the molecule $\sigma_{M}$ as a sum of ionization cross sections of 
the isolated atoms $\sigma_{\alpha}$ weighted by $n_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We classified the molecular targets of our interest in three families: 
CH, CHN and DNA, as in Table~\ref{tab:families}.

\begin{table}[H]
\begin{center}
\begin{tabular}{|p{0.06\textwidth}|p{0.6\textwidth}|}
\hline
\multirow{2}{*}{CH} 
      & CH$_4$ (methane), C$_2$H$_2$ (acetylene), C$_2$H$_4$ (ethene), \\
      & C$_2$H$_6$ (ethane), C$_6$H$_6$ (benzene) \\
\hline
\multirow{2}{*}{CHN} 
      & C$_5$H$_5$N (pyridine), C$_4$H$_4$N$_2$ (pyrimidine), \\ 
      & C$_2$H$_7$N (dimenthylamine), CH$_5$N (monomethylamine) \\
\hline
\multirow{4}{*}{DNA} 
      & C$_5$H$_5$N$_5$ (adenine), C$_4$H$_5$N$_3$O (cytosine), \\
      & C$_5$H$_5$N$_5$O (guanine), C$_5$H$_6$N$_2$O$_2$ (thymine), \\
      & C$_4$H$_4$N$_2$O$_2$ (uracil), C$_4$H$_8$O (THF), \\
      & C$_5$H$_{10}$O$_5$P (DNA backbone), 
        C$_{20}$H$_{27}$N$_7$O$_{13}$P$_2$ 
(dry DNA) \\
\hline
\end{tabular}
\caption{Molecular targets studied in this work, classified in three 
families.}
\label{tab:families}
\end{center}
\end{table}

In Fig.~\ref{fig:crossDNA_1}, we report the total ionization cross 
sections by the impact of multicharged ions for adenine, cytosine, 
guanine and thymine. For adenine, the agreement with the experimental 
data available~\cite{iriki2011} is very good. To the best of our 
knowledge, there are not experimental data on ion--collision ionization 
for the rest of the molecules. We included electron impact 
measurements~\cite{rahman2016} with the corresponding equivelocity 
conversion for incident energies higher than 300~eV. In this region, 
the proton and electron cross section should converge. Although the 
electron impact measurements are above our findings for all the 
molecular targets, it is worth stating that our results agree very well 
with other electron impact theoretical 
predictions~\cite{mozejko2003,tan2018}.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig2.eps}
\includegraphics[width=0.9\textwidth]{figuras/adn1.eps}
\caption{Reduced CDW ionization cross section by impact of multicharged 
ions. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} for proton 
impact and $\square$~\cite{rahman2016} for electron impact with 
equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}
